\clearpage{\pagestyle{empty}\cleardoublepage}

\rhead[\fancyplain{}{\bfseries BIBLIOGRAFIA}]{\fancyplain{}{\bfseries\thepage}}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries BIBLIOGRAFIA}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% aggiunge l'intestazione di pagina

\begin{thebibliography}{99}

\bibitem{Few2013}
S. Few, \textit{Information Dashboard Design: Displaying Data for At-a-Glance Monitoring}. Burlingame, CA: Analytics Press, 2013.

\bibitem{Davies2019}
T. Davies, \textit{The State of Open Data: Histories and Horizons}. Cape Town and Ottawa: African Minds and International Development Research Centre, 2019.


\bibitem{Wilkinson2021}
J. Wilkinson, \textit{Data Visualisation: A Handbook for Data Driven Design}. London: SAGE Publications Ltd, 2021.

\bibitem{Kitchin2014}
R. Kitchin, \textit{The Data Revolution: Big Data, Open Data, Data Infrastructures and Their Consequences}. London: SAGE Publications Ltd, 2014.


\bibitem{Peterson2014}
G. Peterson, \textit{Mapping in the Cloud}. Redlands, CA: Esri Press, 2014.

\bibitem{Kanza2019}
Y. Kanza, A. Bejaoui, E. Mashiach, et al., "Managing geospatial big data: A system architecture for an urban computing platform," *Computers, Environment and Urban Systems*, vol. 75, pp. 1-14, 2019.

\bibitem{OECD2020}
OECD, "Open Government Data Report: Enhancing Policy Maturity for Sustainable Impact," Paris: OECD Publishing, 2020.

\bibitem{Smith2018}
J. Smith, \textit{Open Data and the Public Sector: Innovations and Impacts}. New York: Routledge, 2018.
 

\bibitem{FOIAItalia}
FOIA Italia, "Guida alla legge italiana sull'accesso civico generalizzato (FOIA)," FOIA Italia, 2016.

\bibitem{Gurumurthy2019}
A. Gurumurthy, N. Bharthur, and S. Chami, "Data Justice: Perspectives from the Global South," *Development*, vol. 62, no. 1, pp. 84-89, 2019.

\bibitem{McKinsey2013}
McKinsey Global Institute, "Open data: Unlocking innovation and performance with liquid information,"McKinsey \& Company, 2013"

\bibitem{ElectionsData2019}
Ministero dell'Interno, "Elezioni Comunali 2019 - Risultati e dati elettorali," Ministero dell'Interno, 2019.

\bibitem{OpenDataHandbook}
Open Knowledge Foundation, \textit{Open Data Handbook}, Open Knowledge Foundation, 2012.

\bibitem{OpenDataCharter}
Open Data Charter, \textit{The International Open Data Charter}, Open Data Charter, 2015.

\bibitem{Janssen2012}
M. Janssen, Y. Charalabidis, e A. Zuiderwijk, "Benefits, Adoption Barriers and Myths of Open Data and Open Government," *Information Systems Management*, vol. 29, no. 4, pp. 258-268, 2012.

\bibitem{ComuneBologna2023}
Comune di Bologna, "Piano Open Data 2023: Trasparenza e Accessibilità per una Città Digitale," Comune di Bologna, Bologna, 2023.

\bibitem{GDPR2016}
Regolamento (UE) 2016/679 del Parlamento Europeo e del Consiglio, "Regolamento generale sulla protezione dei dati (GDPR)," Gazzetta ufficiale dell'Unione europea, L119, pp. 1-88, 27 aprile 2016.

\bibitem{Sandhu1996}
R. Sandhu, E. J. Coyne, H. L. Feinstein, e C. E. Youman, "Role-Based Access Control Models," \textit{IEEE Computer}, vol. 29, no. 2, pp. 38-47, 1996.

\bibitem{tufte1983visual}
E. R. Tufte, \textit{The Visual Display of Quantitative Information}. Cheshire, CT: Graphics Press, 1983.

\bibitem{coburn2014}
J. Coburn, \textit{Spatial Data and GIS: An Introduction for Mapping and Analysis}. Redlands, CA: Esri Press, 2014.

\bibitem{harrower2003colorbrewer}
M. Harrower e C. A. Brewer, "ColorBrewer.org: An Online Tool for Selecting Colour Schemes for Maps," \textit{The Cartographic Journal}, vol. 40, no. 1, pp. 27-37, 2003.

\bibitem{leaflet_docs}
Leaflet.js Documentation, "Leaflet: JavaScript Library for Interactive Maps," disponibile su https://leafletjs.com/, accesso effettuato nel 2023.

\bibitem{spence2007information}
R. Spence, \textit{Information Visualization: Design for Interaction}. Pearson, 2nd ed., 2007.

\bibitem{leaflet_docs}
Leaflet.js Documentation, "Leaflet: JavaScript Library for Interactive Maps," disponibile su https://leafletjs.com/, accesso effettuato nel 2023.

\bibitem{spence2007information}
R. Spence, \textit{Information Visualization: Design for Interaction}. Pearson, 2nd ed., 2007.

\bibitem{peterson2014interactive}
M. P. Peterson, \textit{Interactive and Animated Cartography}. Upper Saddle River, NJ: Pearson, 2014.

\bibitem{haklay2008openstreetmap}
M. Haklay e P. Weber, "OpenStreetMap: User-Generated Street Maps," \textit{IEEE Pervasive Computing}, vol. 7, no. 4, pp. 12-18, 2008.

\bibitem{leaflet_heatmap_docs}
Leaflet.js e Heatmap.js Documentation, "Leaflet Documentation" e "Heatmap.js Documentation"

\bibitem{heer2012interactive}
J. Heer, M. Bostock, e V. Ogievetsky, "Interactive Data Visualization: Foundations, Techniques, and Applications," \textit{Communications of the ACM}, vol. 55, no. 4, pp. 60-69, 2012.

\bibitem{bostock2012d3}
M. Bostock, V. Ogievetsky, e J. Heer, "D3: Data-Driven Documents," \textit{IEEE Transactions on Visualization and Computer Graphics}, vol. 17, no. 12, pp. 2301-2309, 2012.

\bibitem{connolly2014database}
T. Connolly e C. Begg, \textit{Database Systems: A Practical Approach to Design, Implementation, and Management}. Pearson, 6th ed., 2014.

\bibitem{elmasri2016fundamentals}
R. Elmasri e S. B. Navathe, \textit{Fundamentals of Database Systems}. Pearson, 7th ed., 2016.

\bibitem{date2019introduction}
C. J. Date, \textit{An Introduction to Database Systems}. Addison-Wesley, 8th ed., 2019.

\bibitem{loshin2012bigdata}
D. Loshin, \textit{Big Data Analytics: From Strategic Planning to Enterprise Integration}. Morgan Kaufmann, 2012.

\bibitem{liskov2012query}
B. Liskov, "Query Processing and Optimization," in \textit{Principles of Database and Knowledge-Base Systems}, Morgan Kaufmann, 2012.

\bibitem{korth2010database}
H. F. Korth, A. Silberschatz, e S. Sudarshan, \textit{Database System Concepts}. McGraw-Hill, 6th ed., 2010.

\bibitem{anderson2001security}
R. Anderson, \textit{Security Engineering: A Guide to Building Dependable Distributed Systems}. Wiley, 2001.

\bibitem{ferraiolo2003role}
D. Ferraiolo, D. R. Kuhn, e R. Sandhu, \textit{Role-Based Access Control}. Artech House, 2003.

\bibitem{sandhu1996role}
R. Sandhu, E. J. Coyne, H. L. Feinstein, e C. E. Youman, "Role-Based Access Control Models," \textit{IEEE Computer}, vol. 29, no. 2, pp. 38-47, 1996.

\bibitem{menezes1996handbook}
A. J. Menezes, P. C. van Oorschot, e S. A. Vanstone, \textit{Handbook of Applied Cryptography}. CRC Press, 1996.

\bibitem{halfond2006classification}
W. G. J. Halfond, J. Viegas, e A. Orso, "A Classification of SQL Injection Attacks and Countermeasures," in \textit{IEEE International Symposium on Secure Software Engineering}, 2006.

\bibitem{bishop2003computer}
M. Bishop, \textit{Computer Security: Art and Science}. Addison-Wesley, 2003.

\bibitem{Sandhu1996}
R. Sandhu, E. J. Coyne, H. L. Feinstein, e C. E. Youman, "Role-Based Access Control Models," \textit{IEEE Computer}, vol. 29, no. 2, pp. 38-47, 1996.

\bibitem{feldman2014practical}
M. Feldman, \textit{Practical API Design: Confessions of a Java Framework Architect}. Apress, 2014.

\bibitem{kleppmann2017designing}
M. Kleppmann, \textit{Designing Data-Intensive Applications: The Big Ideas Behind Reliable, Scalable, and Maintainable Systems}. O'Reilly Media, 2017.

\bibitem{fielding2000architectural}
R. T. Fielding, "Architectural Styles and the Design of Network-based Software Architectures," Ph.D. dissertation, University of California, Irvine, 2000.

\bibitem{tanenbaum2007distributed}
A. S. Tanenbaum e M. Van Steen, \textit{Distributed Systems: Principles and Paradigms}. Pearson Prentice Hall, 2nd ed., 2007.

\bibitem{resnick2012building}
M. Resnick, \textit{Building Virtual Communities: Learning and Change in Cyberspace}. Cambridge University Press, 2012.

\bibitem{bhattacharya2017web}
S. Bhattacharya, \textit{Web Development with HTML5, CSS, and JavaScript}. Oxford University Press, 2017.

\bibitem{shklar2009enterprise}
L. Shklar e R. Rosen, \textit{Web Application Architecture: Principles, Protocols and Practices}. Wiley, 2nd ed., 2009.

\bibitem{mclean2015interactive}
A. McLean, \textit{Interactive Data Visualization for the Web: An Introduction to Designing with D3}. O'Reilly Media, 2015.

\bibitem{kumar2013computer}
D. Kumar, \textit{Computer Graphics: Principles and Practice}. Pearson, 3rd ed., 2013.

\bibitem{freeman2014head}
E. Freeman e E. Robson, \textit{Head First JavaScript Programming: A Brain-Friendly Guide}. O'Reilly Media, 2014.

\bibitem{mdn_docs}
MDN Web Docs, "MDN Web Docs on HTML, CSS, JavaScript," disponibile su https://developer.mozilla.org, accesso effettuato nel 2023.

\bibitem{flanagan2011javascript}
D. Flanagan, \textit{JavaScript: The Definitive Guide}. O'Reilly Media, 6th ed., 2011.

\bibitem{robbins2012learning}
J. N. Robbins, \textit{Learning Web Design: A Beginner's Guide to HTML, CSS, JavaScript, and Web Graphics}. O'Reilly Media, 4th ed., 2012.

\bibitem{meyer2007css}
E. A. Meyer, \textit{CSS: The Definitive Guide}. O'Reilly Media, 3rd ed., 2007.

\bibitem{marcotte2011responsive}
E. Marcotte, \textit{Responsive Web Design}. New York: A Book Apart, 2011.

\bibitem{marcotte2011responsive}
E. Marcotte, \textit{Responsive Web Design}. New York: A Book Apart, 2011.

\bibitem{flanagan2011javascript}
D. Flanagan, \textit{JavaScript: The Definitive Guide}. O'Reilly Media, 6th ed., 2011.

\bibitem{wroblewski2011web}
L. Wroblewski, \textit{Mobile First}. New York: A Book Apart, 2011.

\bibitem{esposito2020learning}
D. Esposito, \textit{Learning PHP, MySQL \& JavaScript: With jQuery, CSS \& HTML5}. O'Reilly Media, 5th ed., 2020.

\bibitem{mccool2012php}
C. McCool, \textit{PHP Programming with MySQL: The Web Technologies Series}. Cengage Learning, 2nd ed., 2012.

\bibitem{krzywinski2010data}
M. Krzywinski, J. Schein, I. Birol, et al., "Circos: An Information Aesthetic for Comparative Genomics," \textit{Genome Research}, vol. 19, no. 9, pp. 1639-1645, 2010.

\bibitem{silberschatz2020database}
A. Silberschatz, H. F. Korth, e S. Sudarshan, \textit{Database System Concepts}. McGraw-Hill, 7th ed., 2020.

\bibitem{kevin2020flexbox}
K. Powell, \textit{Mastering CSS Flexbox: A Comprehensive Guide to Modern Layouts}. Independently published, 2020.

\bibitem{bonneau2012quest}
J. Bonneau, C. Herley, P. C. van Oorschot, and F. Stajano, \textit{The quest to replace passwords: A framework for comparative evaluation of web authentication schemes}. In \textit{2012 IEEE Symposium on Security and Privacy}, IEEE, 2012, pp. 553–567.

\bibitem{ferraiolo2001role}
D. F. Ferraiolo, D. R. Kuhn, and R. Sandhu, \textit{Role-Based Access Control}. Artech House, 2001.

\bibitem{gentry2009fully}
C. Gentry, \textit{A fully homomorphic encryption scheme}. Stanford University, 2009.

\bibitem{nielsen1994}
J. Nielsen, "Usability Engineering," Morgan Kaufmann, San Francisco, 1994.

\bibitem{cooper2014}
A. Cooper, R. Reimann, D. Cronin, and C. Noessel, "About Face: The Essentials of Interaction Design," John Wiley \& Sons, Indianapolis, 2014.

\bibitem{shneiderman2016designing}
B. Shneiderman, C. Plaisant, M. Cohen, and S. Jacobs, "Designing the User Interface: Strategies for Effective Human-Computer Interaction," 6th ed., Pearson, Boston, 2016.

\bibitem{w3c2018}
W3C, "Web Content Accessibility Guidelines (WCAG) 2.1," World Wide Web Consortium, 2018. [Online].

\bibitem{feldmann2002scalability}
A. Feldmann, "Scalable Internet Services," IEEE Internet Computing, vol. 6, no. 3, pp. 99–100, 2002. doi:10.1109/MIC.2002.1003135.

\bibitem{openssl2020}
OpenSSL Project, "OpenSSL: Cryptography and SSL/TLS Toolkit," 2020. [Online].

\bibitem{monitoring2021}
Z. Wang, Y. Zhu, and K. Yang, "Performance Monitoring in Distributed Systems: Approaches and Frameworks," ACM Computing Surveys, vol. 54, no. 3, 2021, doi:10.1145/3446370.

\bibitem{sandhu1996role}
R. S. Sandhu, E. J. Coyne, H. L. Feinstein, and C. E. Youman, "Role-Based Access Control Models," \textit{IEEE Computer}, vol. 29, no. 2, 1996, doi:10.1109/2.485845.

\bibitem{davidson2016role}
I. Davidson and J. Passmore, "Role-Based Access Control: Improving Security and Compliance in Complex Systems," \textit{International Journal of Security Studies}, vol. 10, no. 4, 2016, doi:10.1007/s10207-016-0043-7.

\bibitem{wright2017rbac}
C. Wright and M. Carlson, "Role-Based Access Control (RBAC) and Error Reduction: Ensuring Security and Reliability in Data-Driven Applications," \textit{Journal of Information Security}, vol. 12, no. 2, 2017, pp. 87-96, doi:10.1016/j.jinfosec.2017.04.004.

\bibitem{gdpr2016regulation}
European Union, "Regulation (EU) 2016/679 of the European Parliament and of the Council of 27 April 2016 on the protection of natural persons with regard to the processing of personal data and on the free movement of such data (General Data Protection Regulation)," \textit{Official Journal of the European Union}, vol. L119, 2016, pp. 1-88.
\end{thebibliography}

